Communication complexity: Chapter 3 \\ Distributional communication complexity

### 2019-01-25 CS860-CommunicationComplexity/assignments/Chapter3.tex
Add files via upload

@@

@@

@@

@@

@@

@@

\begin{theorem}
There exists a distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ for which 
$
\disc_\mu(\Eq) \ge \frac18
$
and so
\[
D^\mu_\epsilon(\Eq) \ge 1
\]
for every $\epsilon \le \frac38$.
\end{theorem}

@@

\begin{theorem}
For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ and every constant $\epsilon > 0$,
\[
D^\mu_{1/4}(\Eq) = O(1).
\]
\end{theorem}

>>>>

\begin{proof}
Consider the trivial protocol which just outputs $1$ on every pair of inputs. Then $\Pr[\pi(x,y)\neq \Eq(x,y)]$ is precisely the probability that $x=y$. Hence
$$\Pr[\pi(x,y)\neq \Eq(x,y)] = \Pr[x=y] = (\frac{1}{2})^n$$
since each bit of $x$ and $y$ is chosen independently and uniformly at random.
\end{proof}

@@

\begin{proof}
$\chi(f) \geq \discu(f)^{-1}$ if and only if
$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right| \geq |\calX||\calY|.$$

Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  = \sum_{R\in \mathcal R} \max_{A \subseteq \calX, B \subseteq \calY}\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right| \geq\sum_{R \in \mathcal{R}} \left| \sum_{(x,y) \in R} \big(M^{\pm}_f\big)_{x,y} \right|= \sum_{R \in \mathcal R} |R|$$
using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
$$\sum_{R \in \mathcal R} |R| = |\calX||\calY|$$
and hence the result holds.
\end{proof}

@@

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mathrm{unif}}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
Let $\{R_1, \dots, R_t\}$ be set of $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr((x,y) \in R_i)\Pr(\pi(x,y) = f(x,y) \mid(x,y) \in R_i) = \sum_{i=1}^t \frac{|R_i|}{|\calX||\calY|}\frac{|R_i\cap f^{-1}(\pi(x,y))|}{|R_i|}.$$
Symmetrically
$$\Pr[\pi(x,y) \neq f(x,y)] = \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i \cap f^{-1}(\pi(x,y)\oplus 1)|.$$
Thus we have
\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
&= \frac{1}{|\calX||\calY|} \sum_{i=1}^t|R_i\cap f^{-1}(\pi(x,y))| -|R_i \cap f^{-1}(\pi(x,y)\oplus 1)| \\
&\leq \frac{1}{|\calX||\calY|} \sum_{i=1}^t  \left| \sum_{(x,y) \in R_i} \big(M^{\pm}_f\big)_{x,y} \right|\\
&\leq t\cdot\discu(f).\end{align*}
Since $\log_2(t) = D_\epsilon^{\mathrm{unif}}(f)$ this completes the proof.
\end{proof}

@@

\begin{proof}
For convenience, define $M:= M^{\pm}_\IP$ and let $\mathbf{1}_S$ denote characteristic vector of a set $S\subseteq [n]$. Now consider an entry $x,y$ of $M^TM$:
$$M^TM_{x,y} = \sum_{z\in\{0,1\}^n} (-1)^{x\cdot z}(-1)^{y\cdot z} = \sum_{z \in \{0,1\}^n}(-1)^z\cdot(x+y) = \begin{cases}
(-1)2^{n-1} + (1)2^{n-1}, &\text{if $x\neq y$} \\
\sum_{z \in \{0,1\}^n}1, &\text{otherwise}.
\end{cases}$$
Hence $$M^TM_{x,y} = \begin{cases}
0, &\text{if $x\neq y$} \\
2^n, &\text{otherwise}
\end{cases}$$
i.e. $M^TM = 2^nI$.

Consider a rectangle $R = A\times B$. Then we have
\begin{align*}
\left| \sum_{x \in A, y \in B} \big(M^{\pm}_f\big)_{x,y} \right|  &= \left| \mathbf{1}_A^T M \mathbf{1}_B\right | \\
&\leq ||\mathbf{1}_A|| \cdot || M\mathbf{1}_B|| &\text{by Cauchy-Schwarz} \\
&= \sqrt{|A|}\cdot\sqrt{\mathbf{1}^T_BM^TM\mathbf{1}_B} \\
&= 2^{n/2}\sqrt{|A|}\sqrt{\mathbf{1}^T_B\mathbf{1}_B} \\
&=2^{n/2}\sqrt{|A|}\sqrt{|B|}.
\end{align*}
Therefore
$$\discu(\IP) \leq \frac{1}{2^n2^n}\max_{A \subseteq \calX, B \subseteq \calY} 2^{n/2}\sqrt{|A|}\sqrt{|B|} \leq 2^{-2n}2^n2^\frac{n}{2} = 2^{-n/2}.$$
\end{proof}

@@

\begin{proof}

$\chi(f) \geq \disc_\mu(f)^{-1}$ if and only if
$$\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left|\mu((A\times B)\cap f^{-1}(0) -\mu((A\times B)\cap f^{-1}(1)) \right| \geq 1.$$

Consider a partition of the input space into $f$-monochromatic rectangles, $\mathcal R$, such that $|\mathcal R| = \chi(f)$. Then
\begin{align*}
&\chi(f)\max_{A \subseteq \calX, B \subseteq \calY}\left|\mu((A\times B)\cap f^{-1}(0) -\mu((A\times B)\cap f^{-1}(1)) \right| \\
& = \sum_{R\in \mathcal R}\max_{A \subseteq \calX, B \subseteq \calY}\left|\mu((A\times B)\cap f^{-1}(0) -\mu((A\times B)\cap f^{-1}(1)) \right|\\ 
&\geq \sum_{R \in \mathcal{R}}\left|\mu(R)\cap f^{-1}(0) -\mu(R\cap f^{-1}(1)) \right|\\ 
&= \sum_{R \in \mathcal{R}} \mu(R)
\end{align*}
using that each $R$ is $f$-monochromatic to obtain the last equality. Since $\mathcal R$ partitions the input space,
$$\sum_{R \in \mathcal R} \mu(R)=1$$
and hence the result holds.
\end{proof}

@@

\begin{proof}
Let $\pi$ be the protocol attaining $D_\epsilon^{\mu}(f)$. Since $\Pr[\pi(x,y) \neq f(x,y)] \leq \epsilon$ and $\Pr[\pi(x,y) = f(x,y)] \geq 1-\epsilon$,
$$1-2\epsilon \leq \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)].$$
Let $\{R_1, \dots, R_t\}$ be set of $f$-monochromatic rectangles corresponding to the path from the root of $\pi$ to each respective leaf. Then
$$\Pr[\pi(x,y) = f(x,y)] = \sum_{i=1}^t \Pr[(x,y) \in R_i \text{ and }\pi(x,y) = f(x,y) ] = \sum_{i=1}^t \mu(R_i \cap f^{-1}(\pi(x,y))).$$
Symmetrically
$$\Pr[\pi(x,y) \neq f(x,y)] = \sum_{i=1}^t\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1)).$$
Thus we have
\begin{align*}1-2\epsilon &= \Pr[\pi(x,y) = f(x,y)] - \Pr[\pi(x,y) \neq f(x,y)] \\
&=\sum_{i=1}^t|\mu(R_i\cap f^{-1}(\pi(x,y))) -\mu(R_i \cap f^{-1}(\pi(x,y)\oplus 1))| \\
&\leq \sum_{i=1}^t \disc_\mu(f)\\
&= t\cdot\disc_\mu(f).\end{align*}
Since $\log_2(t) = D_\epsilon^{\mu}(f)$ this completes the proof.
\end{proof}

@@

\begin{theorem}
There exists a distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$ for which 
$
\disc_\mu(\Eq) \leq \frac18 \frac{2^n}{(2^n-1)}.
$
and so
\[
D^\mu_\epsilon(\Eq) \ge 1
\]
for every $\epsilon \le \frac38$.
\end{theorem}

@@

\begin{theorem}
For every distribution $\mu$ over $\{0,1\}^n \times \{0,1\}^n$,
\[
D^\mu_{1/4}(\Eq) = O(1).
\]
\end{theorem}
